Gas exchange in large rivers controlled by hydraulic geometry: implications for remotely sensing gas exchange via the SWOT satellite

### 2021-09-10 manuscript_v5.Rmd
first ms push with hydraulic geometry stuff

---
title: "Gas exchange in large rivers controlled by largest turbulent eddies: implications for remotely sensing gas exchange via SWOT"
author: 'CB Brinkerhoff, CJ Gleason, CJ Zappa, PA Raymond, MH Harlan'
date: "`r Sys.Date()`"
bibliography: References.bib
csl: agu_citation_style.csl
output:
  word_document: 
      reference_docx: word-styles-reference.docx
---

@@

##Key Points
- In large rivers, there exists a local equilibrium between the production and dissipation of turbulent kinetic energy at the free surface
- Gas exchange in large rivers is equivalently parameterized across multiple turbulence scales, with the classical Kolmogorov model recovered empirically
- BIKER algorithm exploits this theory, predicting gas exchange velocity and $CO_2$ fluxes from simulated SWOT satellite data and in situ $CO_2$ data

##Keywords
gas exchange, open-channel flow, remote sensing, SWOT, biogeochemistry, ungauged basin

@@

##1 Introduction
Natural systems play a fundamental role in the budgeting and accounting of the global carbon cycle under climate change. Since the publication of @colePlumbingGlobalCarbon2007a, the global river network is recognized to emit substantial amounts of carbon to the atmosphere, in addition to exporting it to the oceans. Current estimates of total carbon dioxide evasion ($FCO_2$) to the atmosphere from the global river network vary from 650-1800 Tg C/yr [@lauerwaldSpatialPatternsCO22015; @raymondGlobalCarbonDioxide2013a] **maybe add Shaoda here**, with 167 Tg-C/yr coming from mountain streams alone [@horgbyUnexpectedLargeEvasion2019]. Despite its incredibly small percentage of the global land surface [0.47%- @raymondGlobalCarbonDioxide2013a], this flux is on par with the total oceanic $CO_2$ uptake rate [@gruberOceanicSinkAnthropogenic2019; @horgbyUnexpectedLargeEvasion2019] and the global forest carbon uptake rate [@panLargePersistentCarbon2011]. It is still relatively poorly constrained but is clearly a critical component of the global carbon cycle. Equation 1 represents this flux given the gas concentration gradient between the water $CO_{2_{water}}$ and the air $CO_{2_{air}}$ and the gas exchange velocity *k*.

@@

The structure of equation 1 necessitates that calculations of this flux are highly sensitive to one's measurements/estimates of *k* [@hallGasExchangeStreams2020]. Even more broadly, fluvial *k* for various dissolved gases are of significant importance to aquatic ecologists modeling stream metabolism [e.g. @bernhardtMetabolicRegimesFlowing2018] and water quality engineers modeling river responses to pollutant loadings [among other things- @chapraSurfaceWaterQualityModeling2008]. Decades of work have focused on elucidating the physical mechanisms behind *k* in oceans and lakes [@wanninkhofAdvancesQuantifyingAirSea2009] and, to a lesser extent, fluvial gas exchange [@hallGasExchangeStreams2020]. Given that *k* should scale with turbulence in a turbulent flow [@hallGasExchangeStreams2020], a generalized model for *k* (equation 2) relies on the characteristic turbulence velocity scale $v_K$, which itself is a function of the characteristic length and time scales [@katulStructureFunctionModel2018]. Extensive field and labratory experiments have confirmed the validity of this model when $v_K$ is set to the Kolmogorov velocity scale [e.g. @lamontEddyCellModel1970; @zappaEnvironmentalTurbulentMixing2007; @lorkeUnifiedScalingRelation2006; @zappaVariationSurfaceTurbulence2003; @moogStreamReaerationNonuniform1999; @wangCoefficientsSmallEddy2015; @vachonRelationshipNearsurfaceTurbulence2010], where the idea is that turbulent kinetic energy (TKE) is produced by the largest, integral-scale turbulent eddies in the flow and then passed down the energy cascade to progressively smaller eddies until eddy motion is stable and TKE is dissipated as heat at the Kolmogorov-scale: the smallest eddies in a turbulent flow. Equation 2 via Kolmogorov-scale eddies can also be derived from both classic thin-film and surface renewal theories for gas exchange [@wangPhysicallyBasedScaling2021].
$$\mathbf{(2)} k\sim Sc^{-n}v_K$$

While this model works reasonably well in non-fluvial environments, there is considerable uncertainity in how it applies to fluvial systems. Specifically, the influence of the riverbed and its large roughness elements in small streams (i.e. bedforms, logjams, sediment out of suspension) complicates scaling fluvial *k* and might lead to substantially higher *k* in streams with whitewater [e.g. @hallAirWaterOxygen2012; @ulsethDistinctAirWater2019a]. Less attention has been paid to the other side of the stream-to-river continumn, which is often conceptualized in relation to *k* as a hybrid condition of both fluvial and non-fluvial hydraulic properties. To date, the handful of existing field studies of large-river *k* have suggested that *k* begins to be notably influenced by wind dynamics, though little else is well-established [@alinPhysicalControlsCarbon2011; @beaulieuControlsGasTransfer2012a; @wangPhysicallyBasedScaling2021]. These mechanistic uncertainties are additionally limited by a large dearth of field-measured fluvial *k*. @wangPhysicallyBasedScaling2021 attempted to address this by simulating *k* in 35 rivers using stream metabolism modeling [@applingOvercomingEquifinalityLeveraging2018] and in situ dissolved oxygen (DO) datasets, finding that equation 2 is valid in their simulated rivers and that *k*~streamflow relationships breakdown in large rivers. However, they did not directly compare against *k* data in extremely small systems [@ulsethDistinctAirWater2019a] and stopped short of parsing out hydraulic explanations for why this happens.

@@

In this context, here we revist the fundamental mechanisms behind gas exchange and turbulence in large rivers using one of the largest available datasets of field-measured river hydraulics [@brinkerhoffReconcilingAtaStationAtManyStations2019] to answer the following question: does *k* behave fundamentally differently in large rivers than in other streams and how does this relate to the classical model of gas evasion in aquatic systems (equation 2)? We exploit the findings from this simple analysis to develop a novel methodology that predicts $k_{600}$ (*k* normalized to a Schmidt number of 600) and its explicit uncertainity solely using SWOT observations. The method requires no in situ inputs of any kind (although in situ data can be ingested and will improve results). We name the RS of $k_{600}$ algorithm BIKER, or the 'Bayesian Inference/Inversion of the $k_{600}$ Evasion Rate' and validate it for 47 SWOT-observable rivers from around the world using hydraulic models to produce SWOT-like data (as SWOT has not yet launched). We also quantify BIKER's sensitivity to the expected SWOT measurement errors on 17 of those rivers: while SWOT data represent a sea change in inland water monitoring, it is expected to have an approximately 10cm error in water surface elevation [@biancamariaSWOTMissionIts2016] as well as river width errors [@frassonExploringFactorsControlling2021]. Finally, we use previously published dissolved $CO_2$ data to represent a hypothetical in situ $CO_2$ sensor and compare the bulk carbon efflux from the 47 rivers as calculated using BIKER and previously published in situ techniques for predicting $k_{600}$.

@@

TKE and gas exchange theory (section 2) are explored using two datasets. First, we use a previously published compilation of field measurements that were originally made to calibrate United States Geological Survey (USGS) streamgauge rating curves and gathered by @brinkerhoffReconcilingAtaStationAtManyStations2019. That dataset contains over 530,000 unique measurements of river channel velocity, width, depth, and discharge from across the continental United States. This is to our knowledge the largest collection of field measurements of river channel hydraulics. Second, we use the dataset from @ulsethDistinctAirWater2019a who measured and/or collected from the literature over 700 measurements of stream hydraulics and $k_{600}$ and is, to our knowledge, the largest collection of field-measured river and stream $k_{600}$. This is also the dataset used to validate our *k* scaling models.

@@

##3 The TKE budget in large rivers and implications for scaling fluvial gas exchange velocity
Turbulence is fundamental to fluvial geomorphology and open-channel flows and has been the subject of extensive research across scientific and engineering disciplines for over a century. Turbulent kinetic energy (TKE) is the energy generated and dissipated by the eddies created from the chaotic nature of turbulent flow. @nakagawaTurbulenceOpenChannel1993 outline the standard energy budget for TKE, at some depth *h*, for a 2-dimensional open-channel flow (equation 3). *G* is the TKE produced, $\epsilon$ is the TKE dissipated, $T_d$ is TKE transported from the riverbed via turbulent diffusion, and $V_d$ is TKE transported via viscous diffusion.
$$\mathbf{(3)} G+T_d+V_d=\epsilon$$

$\epsilon$ at the free surface is frequently described as $\epsilon(H)=\frac{U_*^3}{H}$ where *H* is the flow depth. However in rivers, which exhibit additional depth-scale shear due to downslope flow against the channel geometry, bars, and meanders, energy dissipation occurs much more evenly over the flow depth than this parameterization allows for. It has been shown that, for rivers at least, a more appropriate dissipation model considers stream power (per unit weight water: $\epsilon_D = gSU$), which represents the energy dissipated by the total frictional resistance of the river reach's geometry [@moogStreamReaerationNonuniform1999]. This is often called 'energy dissipation via form-drag' in the literature, and is generally used to upscale *k* in rivers and streams [i.e. @raymondScalingGasTransfer2012a; @ulsethDistinctAirWater2019a]. Both @raymondScalingGasTransfer2012a and @moogStreamReaerationNonuniform1999 confirmed that $\epsilon_D$ is far greater than $\epsilon$ in open-channel flows, therefore generating the majority of fluvial TKE dissipation.

We express *G* at the free surface in two different ways (equation 4) following @nakagawaTurbulenceOpenChannel1993 on the left and our derivation on the right. Text S1 details the necessary algebra to arrive at both expressions of *G*. Looking at equation 4, it is readily apparent that $\frac{R_h}{H}$ will cancel out if they are equal. This occurs in most large and wide rivers and is a common assumption in hydraulic modeling. In this scenario, the following three statements must be true: 1) $G \approx gSU \approx \epsilon_D \approx \epsilon$, 2) $\epsilon(H) \neq \frac{U_*^3}{H}$, and 3) $T_d \approx 0$. More intuitively, if this "large river" condition is met then additonal TKE transported from the riverbed is functionally zero and the TKE budget at the free surface is just stream power (per unit weight water). This implies a local equilibrium between TKE produced and TKE dissipated at the free surface. 
$$\mathbf{(4)} G=\frac{UU_*^2}{H}=gSU\frac{R_h}{H} $$

We test this hypothesis using the data from @brinkerhoffReconcilingAtaStationAtManyStations2019 and @ulsethDistinctAirWater2019a by plotting boxplots of $\frac{R_h}{H}$ ratios for over 531,000 different measurements of river hydraulics versus $\epsilon_D - G$, which is equivalent to $T_d$ in equation 3 assuming that $V_d$ is neglibile away from the wall [@nakagawaTurbulenceOpenChannel1993]. *G* is calculated as $\frac{UU_*^2}{H}$. In order to determine when $R_h=H$, we round all ratios to 3 significant figures and set all ratios greater than or equal to 0.995 as $R_h=H$. If our hypothesis holds, then $\epsilon_D - G$ should be functionally zero only when the 'large river' condition is met.

![Figure 2: Rh/H versus eD-G at the free surface for our two datasets of field-measured river channel hydraulics (see Appendix A for variable descriptions). eD-G is functionally zero only when a river is sufficiently large. This indicates that a local equilibrium of TKE is acheived and energy generated by large-scale eddies is nearly equal to energy dissipated by the smallest scale eddies. In this figure, Rh/H bins contain between 573 and 138,307 measurements.](cache/k600_theory/turbulence.jpg)

Figure 2, which plots this test, shows that $\epsilon_D-G$ is nearly two orders of magnitude less when $R_h=H$ than it is when $\frac{R_h}{H}$ is between 90-99% (median value across both datasets of $10^{-5.39}$ versus $10^{-3.83} \frac{J}{kg*s}$, respectively). When the $\frac{R_h}{H}$ ratio is very small (0-9%), this rises to a median value of $10^{-2.56} \frac{J}{kg*s}$, or nearly three orders of magnitude larger. The $\epsilon_D - G$ values when $R_h=H$ (which are generally $10^{-5}$ to $10^{-8} \frac{J}{kg*s}$) are functionally zero. This means that a local equilibrium of free-surface TKE is achieved, where energy produced via the largest scale (turbulent) eddies is equivalent to energy dissipated via the smallest scale (Kolmorogov) eddies. This idea of a local equilibrium is certainly not new, however this is to our knowledge the first confirmation that it happens naturally in most large rivers and that is arises algebraically from the TKE budget in open-channel flows. From here on we refer to the $R_h=H$ river scenario as the 'large river' condition.

This finding has a significant impact on how we model *k* in large rivers. Because the local equilibrium acheived during the 'large river condition' necessitates that $\epsilon$ is equivalent to *G* (Figure 2), we can presumably set $v_K$ in equation 2 equal to the largest ('integral-scale') eddies in the system, which are much more easily remotely-sensible than the Kolmogorov velocity scale. Otherwise, *k* should continue to scale only using the Kolmogorov-scale (which is consistent across all turbulent flows by its definition).

To test this hypothesis, we fit the following three models to the @ulsethDistinctAirWater2019a dataset of *k* measurements normalized to a Schmidt number of 600: $k_{600}=\beta U_*$, $k_{600}=\alpha \epsilon^{1/4}$, and $k_{600}=a (\epsilon_D)^b$ under two scenarios: $R_h=H$ and $R_h =/= H$. Note that because all data are normalized to a Schmidt number of 600, both the Schmidt numbers and viscosity terms are constant and are implicit in the statistical coefficients. The model structures (and the fit parameters) are detailed in Table 1. The third model tested (rows 5 and 6 of Table 1), was fit to see which slope coefficient is retrieved empirically by fitting the generalized scaling function to the data. Therfore, the $r2$ for these models is reflective of the log-transformed $k_{600}$ (unlike the other two models). This was the approach used by @ulsethDistinctAirWater2019a, who observed slopes steeper than 1/4 in both regimes of their breakpointed regression model.

![Figure 3: Six k600 predictive models tested on the dataset from Ulseth et al. (2019). a-c are fit to the 'large river condition' while d-f are not. Red coefficients of determination refer to log-tranformed quantities while blue coefficients of determination refer to the natural quantities.](cache/k600_theory/k600Plot.jpg)

The results of this test are presented as Figure 3 and Table 1, where our hypothesis is confirmed: both the $U_*$ and $\epsilon^{1/4}$ models explain ~70% of variance in *k* under the 'large river' condition while the $U_*$ model does not work when this condition is not met ($r^2$:`r round(models_ustar[2,]$r2,2)` versus $r^2$=`r round(models_ustar[1,]$r2,2)`). In fact, Figures 3a and 3b are nearly visually identical in their scaling dynamics, while Figures 3d and 3e are very different, lending further evidence that these two scaling models are synonmous only when the 'large river' condition is met. This is in line with @wangPhysicallyBasedScaling2021, who found similar predictive performance and parameter values for these two models on 35 simulated riverswhich they assumed were large enough that $R_h \approx H$. Further, the 1/4 slope parameter is statistically recovered in the 'large river' condition (row 3, Table 1: slope = `r round(models_ulseth[2,]$slope,2)`). We note that the large residuals in Figure 3c are a byproduct of the model existing in log-space and that panels 3b and 3c are functionally identical models (see Table 1 for their nearly identical intercept and slope parameters). The top row of Figure 3 confirms that under the 'large river condition' *k* can be scaled using either integral or Kolmogorov turbulence scales and that they are synonymous. This opens the door for remotely sensing *k* using almost entirely remotely-sensed parameters. This is elaborated on in section 4.

*Table 1: Model performance and parameter values for the six models tested*

|**Figure 3 Panel**|**River Regime**|**Model Structure**|**r2**|**Intercept Parameter**|**Slope Parameter**|
|---------------------|------------------|-------------------|------|---------------|---------------|------------------|
|3a|$R_h=H$|$k_{600}=\beta U_*$|`r round(models_ustar[2,]$r2, 2)`|`r signif(models_ustar[2,]$int, 3)`|`r round(models_ustar[2,]$slope, 2)`|
|3d|$R_h=/=H$|$k_{600}=\beta U_*$|`r round(models_ustar[1,]$r2, 2)`|`r signif(models_ustar[1,]$int, 3)`|`r round(models_ustar[1,]$slope, 2)`|
|3b|$R_h=H$|$k_{600}=\alpha(\epsilon_D)^{1/4}$|`r round(models_eD4[2,]$r2, 2)`|`r signif(models_eD4[2,]$int, 3)`|`r round(models_eD4[2,]$slope, 2)`|
|3e|$R_h=/=H$|$k_{600}=\alpha(\epsilon_D)^{1/4}$|`r round(models_eD4[1,]$r2, 2)`|`r signif(models_eD4[1,]$int, 3)`|`r round(models_eD4[1,]$slope, 2)`|
|3c|$R_h=H$|$k_{600}=a(\epsilon_D)^{b}$|`r round(models_ulseth[2,]$r2, 2)`|`r signif(models_ulseth[2,]$int, 3)`|`r round(models_ulseth[2,]$slope, 2)`|
|3f|$R_h=/=H$|$k_{600}=a(\epsilon_D)^{b}$|`r round(models_ulseth[1,]$r2, 2)`|`r signif(models_ulseth[1,]$int, 3)`|`r round(models_ulseth[1,]$slope, 2)`|

The 1/4 exponent is clearly not valid when $R_h=/=H$ (Figure 3e). However, the generalized form of this model explains `r 100*round(models_ulseth[1,]$r2,2)`% of the variation in log-transformed *k* and was deemed a strong scaling model for fluvial *k* when rivers do not meet the 'large river' condition. That model yields a slope of `r round(models_ulseth[1,]$slope,2)` which corroborates other models with slopes steeper than 1/4 in rivers [@wangPhysicallyBasedScaling2021; @ulsethDistinctAirWater2019a; @raymondScalingGasTransfer2012a]. Further theoretical modeling is needed to parse out why fluvial *k* in systems that do not meet the 'large river condition' generally scale at a steeper rate than in non-fluvial systems, and how this relates to the breakpointed model proposed by @ulsethDistinctAirWater2019a. That said, the TKE budget in equation 3 suggests that it is a function of turbulent diffusion of additional TKE from the bed to the free surface.

>>>>

---
title: "Gas exchange in large rivers controlled by hydraulic geometry: implications for remotely sensing gas exchange via the SWOT satellite"
author: 'CB Brinkerhoff, CJ Gleason, CJ Zappa, PA Raymond, MH Harlan'
date: "`r Sys.Date()`"
bibliography: References.bib
csl: agu_citation_style.csl
output:
  word_document: 
      reference_docx: word-styles-reference.docx
---

@@

##Key Points
- River channel efficiency effec the relationship between gas exchange and form-drag dissipation of turbulence
- Large, less efficient channel geometry necessitates that near-surface turbulence is dominated by form-drag rather than bed friction
- BIKER algorithm exploits this theory, predicting gas exchange velocity and $CO_2$ fluxes from simulated SWOT satellite data and in situ $CO_2$ data

##Keywords
gas exchange, fluvial geomorphology, open-channel flow, remote sensing, SWOT, biogeochemistry, ungauged basin

@@

##1 Introduction
Natural systems play a fundamental role in the budgeting and accounting of the global carbon cycle under climate change. Since the publication of @colePlumbingGlobalCarbon2007a, the global river network is recognized to emit substantial amounts of carbon to the atmosphere, in addition to exporting it to the oceans. Current estimates of total carbon dioxide evasion ($FCO_2$) to the atmosphere from the global river network vary from 650-1800 Tg C/yr [@lauerwaldSpatialPatternsCO22015; @raymondGlobalCarbonDioxide2013a] **maybe add Shaoda here**, with 167 Tg-C/yr coming from mountain streams alone [@horgbyUnexpectedLargeEvasion2019]. Despite its incredibly small percentage of the global land surface [0.47%- @raymondGlobalCarbonDioxide2013a], this flux is on par with the total oceanic $CO_2$ uptake rate [@gruberOceanicSinkAnthropogenic2019; @horgbyUnexpectedLargeEvasion2019] and the global forest carbon uptake rate [@panLargePersistentCarbon2011]. It is still relatively poorly constrained but is clearly a critical component of the global carbon cycle. Equation 1 represents this flux given the gas concentration gradient between the water $CO_{2_{water}}$ and the air $CO_{2_{air}}$ and the gas exchange velocity *k*. Consult Appendix A for variable nomenclature used in this study.

@@

The structure of equation 1 necessitates that calculations of this flux are highly sensitive to one's measurements/estimates of *k* [@hallGasExchangeStreams2020]. Even more broadly, fluvial *k* for various dissolved gases are of significant importance to aquatic ecologists modeling stream metabolism [e.g. @bernhardtMetabolicRegimesFlowing2018] and water quality engineers modeling river responses to pollutant loadings [among other things- @chapraSurfaceWaterQualityModeling2008]. Decades of work have focused on elucidating the physical mechanisms behind *k* in oceans and lakes [@wanninkhofAdvancesQuantifyingAirSea2009] and, to a lesser extent, fluvial gas exchange [@hallGasExchangeStreams2020]. Given that *k* should scale with near-surface turbulence in a turbulent flow [@hallGasExchangeStreams2020], extensive field and labratory experiments have converged on the 'small-eddy model' as proposed by @lamontEddyCellModel1970, which scales *k* via the smallest-scale turbulent eddies [e.g. @zappaEnvironmentalTurbulentMixing2007; @lorkeUnifiedScalingRelation2006; @zappaVariationSurfaceTurbulence2003; @moogStreamReaerationNonuniform1999; @wangCoefficientsSmallEddy2015; @vachonRelationshipNearsurfaceTurbulence2010]. This model is provided as equation 2, where $\epsilon_S$ is the dissipation rate of near-surface turbulence and $v$ is the kinematic viscosity and *Sc* is the Schmidt number.
$$\mathbf{(2)} k\sim Sc^{-n}(v\epsilon_S)^{1/4}$$

While this model works reasonably well in non-fluvial environments where turbulence production is largely confined to bottom shear or wind shear [e.g. @lorkeUnifiedScalingRelation2006; @zappaEnvironmentalTurbulentMixing2007], there is considerable uncertainity in how it applies to fluvial systems. First, 'bubble-mediated gas exchange' in whitewater might lead to substantially higher *k* in small mountain streams with very large bedload [@hallAirWaterOxygen2012; @ulsethDistinctAirWater2019a]. Additionally, 'form-drag shear' via river channel banks, meanders, bars, etc. further complicate scaling fluvial *k* [@moogStreamReaerationNonuniform1999]. Those authors proposed modeling fluvial *k* via 'form-drag dissipation' $\epsilon_D$, assuming that total reach stream power is responsible for all turbulence production and therefore dissipation. Numerous authors have since shown that this model ($\epsilon_D=gS\overline{U}$) reasonably predicts *k* in rivers [@raymondScalingGasTransfer2012a; @ulsethDistinctAirWater2019a; @moogStreamReaerationNonuniform1999], though it is not synonmous with equation 2 as $\epsilon_D$ refers to a depth-scale disispation rate and is not necessairily indicative of the near-surface dissipation rate. Likewise, $k\sim\epsilon_D$ usually has a slope much steeper than 1/4.

Relatively less attention has been paid to gas exchange in very large rivers, which is often conceptualized as a hybrid condition of both fluvial and non-fluvial hydraulic properties. To date, the handful of existing field studies of large-river *k* have suggested that *k* begins to be influenced by wind shear, though little else is well-established [@alinPhysicalControlsCarbon2011; @beaulieuControlsGasTransfer2012a; @wangPhysicallyBasedScaling2021]. These mechanistic uncertainties are additionally limited by a large dearth of field-measured fluvial *k*. @wangPhysicallyBasedScaling2021 attempted to address this by simulating *k* in 35 rivers using stream metabolism modeling [@applingOvercomingEquifinalityLeveraging2018] and in situ dissolved oxygen (DO) datasets, finding that equation 2 + $\epsilon$ modeled via the log-law-of-the-wall is valid in their simulated rivers and that *k*~streamflow relationships breakdown in large rivers. However, they did not directly compare against *k* data in extremely small systems [@ulsethDistinctAirWater2019a] and stopped short of parsing out hydraulic explanations for why this happens.

@@

In this context, here we revist the fundamental mechanisms behind gas exchange and turbulence in large rivers to answer the following question: does *k*'s relationship with form-drag fundamentally change in large rivers and how does this relate to one of the classical model's of gas evasion in aquatic systems (equation 2)? We exploit the findings from this simple analysis to develop a novel methodology that predicts $k_{600}$ (*k* normalized to a Schmidt number of 600) and its explicit uncertainity solely using SWOT observations. The method requires no in situ inputs of any kind (although in situ data can be ingested and will improve results). We name the RS of $k_{600}$ algorithm BIKER, or the 'Bayesian Inference/Inversion of the $k_{600}$ Evasion Rate' and validate it for 47 SWOT-observable rivers from around the world using hydraulic models to produce SWOT-like data (as SWOT has not yet launched). We also quantify BIKER's sensitivity to the expected SWOT measurement errors on 17 of those rivers: while SWOT data represent a sea change in inland water monitoring, it is expected to have an approximately 10cm error in water surface elevation [@biancamariaSWOTMissionIts2016] as well as river width errors [@frassonExploringFactorsControlling2021]. Finally, we use previously published dissolved $CO_2$ data to represent a hypothetical in situ $CO_2$ sensor and compare the bulk carbon efflux from the 47 rivers as calculated using BIKER and previously published in situ techniques for predicting $k_{600}$.

@@

Gas exchange theory (section 2) is explored using the dataset from @ulsethDistinctAirWater2019a who measured and/or collected from the literature over 700 measurements of stream hydraulics and $k_{600}$ and is, to our knowledge, the largest collection of field-measured river and stream $k_{600}$. To assess general hydraulic conditions in rivers across geographic contexts, we additionally use a previously published compilation of field hydraulics measurements that were originally made to calibrate United States Geological Survey (USGS) streamgauge rating curves [@brinkerhoffReconcilingAtaStationAtManyStations2019]. That dataset contains over 530,000 unique measurements of river channel velocity, width, depth, and discharge from across the continental United States. This is to our knowledge the largest collection of field measurements of river channel hydraulics. Ultimately, this dataset was filtered down to 171,553 discrete measurements (see Text S1 for our filtering protocol).

@@

##3 Form-drag in large rivers and implications for scaling fluvial gas exchange velocity
Turbulence is fundamental to fluvial geomorphology and open-channel flows and has been the subject of extensive research across scientific and engineering disciplines for over a century. Turbulence in river flows is usually generated by shear forces within the river channel (wind shear's influence is usually limited to only the absolute widest rivers). This within-bank shear comes from the riverbed, as well as 'form-drag' via riverbanks, meanders, bars, suspended sediment, etc. Both @raymondScalingGasTransfer2012a and @moogStreamReaerationNonuniform1999 confirmed that $\epsilon_D$ is far greater than $\epsilon_S$ in open-channel flows, therefore generating the majority of fluvial TKE dissipation.

'Form-drag' shear exists at the depth-scale and therefore is scale-dependent: while small roughness elements near the bed produce form-drag, if they are small relative to the total flow depth they will exert little influence on near-surface dissipation [@moogStreamReaerationNonuniform1999]. This means that, if all other properties of a river flow are held constant but the channel is progressively made wider and shallower, the depth-scale will decrease and form-drag will compose a progressively larger component of near surface dissipation. This concept is illustrated graphically in Figure 2 and represents the geomorphic concept of 'river channel efficiency': all else being constant, a less efficient channel (wide/shallow) will have a longer wetted perimeter and therefore a larger shear surface area than the same flow through a narrow/deep channel [@chowOpenChannelHydraulics1959a]. This means that the inefficient channel will 1) exhibit more form-drag and thus turbulence, and 2) turbulence generated via form-drag will more readily reach the surface. Note that to compare wetted perimeters across many rivers with different channel areas, we use the hydraulic radius $R_h$ (channel area divided by the wetted perimeter) as compared to the mean flow depth $H$. An efficient channel will have a low $\frac{R_h}{H}$ ratio while an inefficient channel will have a high $\frac{R_h}{H}$ ratio that approaches 1.

![Figure 2: Conceptual model relating river channel efficiency to the turbulent dissipation rate. All else being constant, a less efficient channel shape (wide/shallow) will 1) have more form-drag and thus turbulence and 2) turbulence generated via form-drag will more readily reach the surface.](cache/k600_theory/conceptual_model.jpg)

This all suggests that, for incredibly inefficient river channels, $\epsilon_S \approx \epsilon_D$. Put another way, the channel produces so much turbulence from form-drag, and the flow is so shallow relative to its width, that depth-scale dissipation is likely synonmous with near-surface dissipation. This would mean that equation 2 could be equivalently written as $k\sim Sc^{-n}(v\epsilon_D)^{1/4}$ when a channel is sufficiently inefficient. To explore when, if at all, this point is reached we fit models using 701 river $k_{600}$ measurements from @ulsethDistinctAirWater2019a. We defined 'channel inefficiency' using minimum $\frac{R_h}{H}$ thresholds from 0.75 to 0.995 and then we fit the two models outlined in equations 3 and 4. Because all of the data are normalized to Schmidt numbers of 600, we can convienantly fold the viscosity and Schmidt numbers into the model coefficients. The goal of this analysis is to find at what $\frac{R_h}{H}$ threshold does 1) equation 3's b parameter approximate 1/4 and 2) equation 4's goodness-of-fit (via the coefficient of determination $r^2$) reach a suitably high value.
$$\mathbf{(3)} k_{600}=a(\epsilon_D)^{b}$$
$$\mathbf{(4)} k_{600}=\beta(\epsilon_S)^{1/4}$$

The results of this analysis are plotted in Figure 3a, along with channel geometry schematics to highlight the progressive shift from efficient to inefficient river channels. Equation 3's b parameter only begins to approximate 1/4 when $R_h$ is within 1% of the mean flow depth. At this point, the model's goodness-of-fit also reaches a reasonable value of 0.68. This indicates that a river is sufficiently inefficient when its hyradulic radius approximates the mean flow depth, and at that point $\epsilon_S \approx \epsilon_D$. Figures 3b-4d plot the model fits for this final scenario where $\frac{R_h}{H} >= 0.995$ As Figure 3c confirms, the small-eddy model is viable when a channel is so inefficient that dissipation via form-drag approximates the near-surface dissipation rate in equation 2. Figure 4d further confirms this, where the slope of the generalized model (equation 3) equals `r round(models_ulseth[models_ulseth$name == 'Rh=H',]$slope, 2)` which is nearly equivalent to 1/4.

![Figure 3: Model development for predicitng k from form-drag dissipation in rivers. (a) Rh/H thresholds versus fitted model results. Dark blue is the fitted b parameter in equation 3 and dark green is the goodness-of-fit of equation 4 as defined by the coefficient of determination. Once a channel is maximally inefficient, i.e. Rh=H, equation 4 is a viable model for k and equation 3 empirically recovers the theoretical value of 1/4 as outlined in equation 2. (b-d) Fitted model results for inefficient channels (defined as Rh/H >= 0.995) predicting k via the shear velocity U*, equation 4, and equation 3 (respectively). Note that colors align to the same models for subpanels a, c, and d. Further note that the coefficient of determination in subpanel (d) refers to log-transformed residuals and hence is not directly comparable to those in subpanels (b) and (c).](cache/k600_theory/k600Plot.jpg)

Finally, becuase $\epsilon_D$ refers to a depth-scale quantity, we tested whether a more parsimonious model is possible by removing the 'middle-man', i.e. the Kolmogorov velocity scale, and directly predicting *k* from depth-scale turbulence. Figure 3b plots fitted model results for the $k=\beta_1U_*$ model. For this data, $U_*$ explains 70% of variation in $k_{600}$ and, as expected, the fitted model is nearly identical to Figure 4c in scaling dynamics (as confirmed visually and via the goodness-of-fit). This all indicates that depth-scale turbulent production/dissipation dominates the surface dissipation signal in maximally inefficient river channels. These findings build off of @ulsethDistinctAirWater2019a's intial result that channel slope exerts a drammatic influence on *k* scaling dynamics, further suggesting that fluvial *k* is as much a function of hydraulic geometry as the turbulence properties of the flow.

With all of that said, how often is a river channel so inefficient that $R_h \approx H$? To answer this question, we join the 701 measurements from @ulsethDistinctAirWater2019a to 171,553 discrete river hydraulics measurements from across the continetal United States [@brinkerhoffReconcilingAtaStationAtManyStations2019]. We find that `r round(percs_overall[percs_overall$flag_depth == 'Rh=H',]$perc,0)`% of hydraulic radii are within 5% of their mean flow depth. Because these measurements were made exclusively at streamgauges where rivers are somewhat channelized and width variability is decreased, we opted for a slightly more liberal ratio of 5%. Because larger rivers are more likely to have inefficient channels, we then flag the measurements that are SWOT-observable, i.e. width > 100m. We find that while `r round(percs[percs$flag_depth == 'Rh=H',]$perc,0)[1]`% of non-SWOT observable rivers meet this inefficient condition, `r round(percs[percs$flag_depth == 'Rh=H',]$perc,0)[2]`% of SWOT-observable rivers do. This significant difference, and near universality of SWOT-observable rivers being inefficient, suggests that our theoretical findings in Figure 3 should hold in nearly all rivers that SWOT will observe. This opens the door for exploiting our 'inefficient river k model' for remote sensing *k* via the SWOT satellite.

##4 Exploiting the 'inefficient river model' to remotely sense gas exchange velocity via the BIKER algorithm
We have shown that scaling $k_{600}$ via the depth-scale shear-velocity ($U_*$) explains ~70% of the variation in $k_{600}$ in inefficient rivers and that nearly all SWOT-observable rivers are inefficient. Further, because of the structure of $U_*$ (Appendix A) we can easily reduce the 'inefficient river' $k_{600}$ model (Figure 3b) to an equation consisting solely of remotely-sensible river hydraulics and an intial estimate of the channel area. By scaling *k* this way, we significantly reduce potential equifinality issues by having just two unknowns: $k_{600}$ and the median channel area $A_0$. Equifinality refers to an under-constrained mathematical system that has essentially infinite parameter combinations that can produce the same result: there are in essence more unknowns than equations [@garamboisInferenceEffectiveRiver2015]. This problem is experienced by both SWOT RSQ algorithms and in situ tools that concurrently solve for *k* and stream metabolism [@applingOvercomingEquifinalityLeveraging2018; @graceFastProcessingDiel2015; @holtgrieveSimultaneousQuantificationAquatic2010]. In both of these domains, the other unknown parameters are often difficult to estimate (bed roughness and stream metabolism, respectively) while for BIKER, median channel area is relatively easy to approximate from the SWOT-observable river width [@brinkerhoffConstrainingRemoteRiver2020a]. Therefore, we develop the BIKER algorithm, which is informed by the @hagemannBAMBayesianAMHGManning2017a algorithm for ungauged RSQ and further explored in more recent work by @brinkerhoffConstrainingRemoteRiver2020a. These papers conceptualize discharge as a Bayesian remote sensing problem, which we largely follow here to conceptualize $k_{600}$ as a Bayesian remote sensing problem that can be solved using SWOT data.
